\Large Last Writer Slices and Communication Traps

### 2013-11-01 paper.tex
committing a version of the paper with a background that serves the main message.

When programmers reason incorrectly, they permit some interleavings that should
be forbidden.  When such interleavings occur, program executions may fail,
experience performance degradation, or suffer other undesirable behavior.
Preventing such undesirable behavior is a fundamental challenge faced by
software engineers.  An important way to understand program behavior is to
monitor which read and write operations depend on one another.  Memory
operations are dependent if their order determines the values they read or
write.  Inter-thread dependence analysis is especially useful for understanding
how threads interact.

@@

Prior work has used dependence information to reduce the burden of making
reliable, efficient concurrent software.  Bug
detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
analyze inter-thread dependences to find buggy behavior and help programmers
understand their errors.
Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
dependences to find optimization opportunities.  Dependences are also essential
to some program understanding systems~\cite{oshatr} and dynamic checkers for
thread interaction properties~\cite{oshajava,velodrome}.
Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
order of dependent events to help reproduce failures.

@@

Many of the systems mentioned above are especially valuable when they run in
production systems on deployed code.  Unfortunately, tracking dependences is
not only a fundamental part of the value of these systems, but also a
fundamental part of their {\em cost}.  Tracking dependences requires executing
additional monitoring code while the program runs and many of the above systems
have reported overheads too high for use in production (10-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.

@@

Framed by these observations, we demonstrate in this work that an important,
versatile form of inter-thread dependence-tracking analysis has inherent
overheads that are low enough for use in deployed systems.  There are two parts
to our work:

@@

Many systems that support programming and execution models, debugging tools,
and other software engineering tools track dynamic memory dependences during a
program's execution.  In this section we provide context for \ctraps by
discussing some of these systems, focusing on those that address concurrency.
We then briefly discuss debugging with data dependence information, in order to
motivate the role of last writer slices at debugging time.

@@

\subsection{Debugging with Data Dependence Information}

Data dependence information is directly useful during program debugging.  A
wide variety of dynamic program slicing
techniques~\cite{thinslicing,tipslicingsurvey} have been proposed.  These
techniques record a subset (slice) of dynamic program instructions selected
using data and control dependence information.  Dynamic dependence slices
expose the provenance of control- and data-flows in the execution.  Work on
debugging~\cite{whylinechi,whylineicse} in the HCI and Software Engineering
community shows that when programmers debug programs, they benefit from the
ability to directly ask such provenance (``why'') questions of their debugger.

Dynamic dependence slices are especially useful for debugging if they are
available after a failure has occurred in a deployed system. However, slicing
in deployed systems is uncommon, because deployed systems demand low overheads
and slicing can have high overhead.  Our system provides provenance information
that is useful at debugging time by collecting {\em last writer slices} that
help understand inter-thread communication.  We track last writer slices
with overhead low enough for use in production code.  Deployable efficiency
makes it possible to collect and report last writer slices the same way today's
systems collect and report core dumps.

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:debugging}
Prior work has demonstrated that it is useful during debugging to allow
programmers to ask ``why?'' questions~\cite{whylineicse,whylinechi} about the
provenance of program state.  To answer such questions, prior work has mostly
used dependence slices that include intra-thread
dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The overhead of
slicing dependence chains and tracking temporal sequences is often high,
relegating such techniques to working on traces~\cite{whylinechi} or debugging
executions run ``in the lab''.

@@

>>>>

When programmers reason incorrectly, they permit some interleavings that should
be forbidden.  When such interleavings occur, program executions may fail,
experience performance degradation, or suffer other undesirable behavior.
Preventing such undesirable behavior is a fundamental challenge faced by
software engineers.  An important way to understand program behavior is to
monitor which read and write operations depend on one another.  Memory
operations are dependent if their order determines the values they read or
write.  The value of dependence information is made especially clear by its
prevalent use during manual program debugging and in support of program
analysis tools.

Debugging a concurrent program requires understanding how threads interact and
dependence information furnishes that understanding.  Programmers debug
programs more effectively when they can ask questions about the provenance of
values during a program's execution~\cite{slicingsurvey,whylineicse}.
Dependence information helps programmers with the difficult task of identifying
control- and data-flows that influenced a particular data value at a particular
point.  In concurrent programs, data provenance information is vital because
influential control- and data-flows may be in any thread, making the task of
understanding the origin of a value much more complex.

Compounding this complexity, many concurrency bugs manifest as failures rarely.
When a deployed system fails and a programmer receives a bug report,
reproducing the reported failure to collect dependence information may be
difficult or impossible.  It is essential to collect as much information as
possible in deployed systems to reduce the need for time-consuming forensic bug
reproduction.  Unfortunately, prior approaches to tracking dependences in
concurrent code~\cite{raceslicing,tipslicingsurvey} impose high run time
overheads.  These overheads are prohibitively high for use in production.

@@

Dependence information is also an important part of many program analysis
tools.  Prior work has used dependence information to reduce the burden of
making reliable, efficient concurrent software.  Bug
detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
analyze inter-thread dependences to find buggy behavior and help programmers
understand their errors.
Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
dependences to find optimization opportunities.  Dependences are also essential
to some program understanding systems~\cite{oshatr} and dynamic checkers for
thread interaction properties~\cite{oshajava,velodrome}.
Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
order of dependent events to help reproduce failures.

@@

These systems would be especially valuable, if run in production systems on
deployed code.  However, as with dependence tracking on its own, most of these
systems have overheads too high for use in production (5-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.

@@

Framed by these observations, we demonstrate in this work that a 
versatile form of inter-thread dependence-tracking analysis has inherent
overheads that are low enough for use in deployed systems.  There are two parts
to our work:

@@

In this section we discuss debugging with data dependence information, using an
example, in order to motivate collecting last writer slices in deployed
systems.  We then provide context for \ctraps by discussing some more
sophisticated concurrency analyses that rely on the kind of dependence
information we are focusing on in this work.

%Many systems that support programming and execution models, debugging tools,
%and other software engineering tools track dynamic memory dependences during a
%program's execution.

\subsection{Debugging and Data Provenance}

Data dependence information is directly useful during program debugging.  Prior
work has proposed a wide variety of dynamic program slicing
techniques~\cite{thinslicing,tipslicingsurvey}.  These techniques record a
subset (slice) of dynamic program instructions selected using data and control
dependence information.  Dynamic dependence slices expose the provenance of
data in the execution.  Work on debugging~\cite{whylinechi,whylineicse} in the
HCI and Software Engineering community shows that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/BackgroundCoreDumpFail.pdf}
\caption{\label{fig:coreDumpFail}{\bf Ordering violation bug in HTTrack.} Simply examining memory does not provide provenance information that helps understand this bug.}
\end{figure}

Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
plays in debugging, using an example of a real bug from HTTrack, a web caching
tool.  The figure depicts two threads manipulating a data structure that
contains a mutex\_lock.  The program has an ordering violation bug: the program
crashes when Thread 2 tries to acquire the lock before Thread 1 has initialized
it.  Assuming the run time environment supports it, a core dump delivered to
the programmer after such a failure reveals very little about how to fix the
bug.  The lock is uninitizalized, so it may hold {\tt NULL} or it may hold
random bits.  Both scenarios provide no insight into the root cause of the
failure, which is that the code permits a use of the lock before its
initialization.   Comparing the core dump from the failing run to the memory
image from a correct execution -- a likely debugging strategy -- reveals only
that in a correct execution, the lock contains valid mutex state.  This
revalation is of limited use as it does not rule out arbitrary data corruption
as the bug's root cause.

The vital, missing piece to solving this bug is provenance information.
Debugging this bug would be easier if the core dump included information about
the thread and code point that last wrote each variable.  The programmer would
see that no thread had initialized the lock in the failing execution, but that
Thread 1 initialized the lock in non-failing executions.  This information
rules out data corruption as a root cause and guides the programmer to add
synchronization ensuring the correct ordering relationship.

Another facet to debugging this bug is that it manifests as a failure
infrequently and under only real-world conditions.  If the programmer runs an
instrumented or tooled version of the program, the failure may not manifest in
a reasonable amount of time and running it may require unavailable real-world
inputs.  Thus, collecting provenance or other information offline is
time-consuming or impossible.  A better strategy is to continuously collect
that information in deployed systems and package it with core dumps sent to
programmers with bug reports.  The last writer slicing design we propose in
this work espouses this {\em in situ} data collection strategy and does so with
overheads appropriate for production systems.

%Dynamic dependence slices are especially useful for debugging if they are
%available after a failure has occurred in a deployed system. However, slicing
%in deployed systems is uncommon, because deployed systems demand low overheads
%and slicing can have high overhead.  Our system provides provenance information
%that is useful at debugging time by collecting {\em last writer slices} that
%help understand inter-thread communication.  We track last writer slices with
%overhead low enough for use in production code.  Deployable efficiency makes it
%possible to collect and report last writer slices the same way today's systems
%collect and report core dumps.

@@

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:debugging}
The WhyLine debugger~\cite{whylineicse} demonstrated that it is useful to
enable programmers to make data provenance queries during debugging.  To answer
such questions, prior work has mostly used dependence slices that include
intra-thread dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The
overhead of slicing dependence chains and tracking temporal sequences is often
high, relegating such techniques to working on traces~\cite{whylinechi} or
debugging executions run ``in the lab''.

@@

\paragraph{Case Study: HTTrack}
